A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-18 tex/body.tex
more comments by Andreas

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$).
\end{rep}
Names starting with an uppercase letter \begin{rep}(e.g.,
$\const{S}$) \end{rep}are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (for $n \ge 0$)
and a return type $\tau$, written
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$\begin{rep},
which collapses to $\const{f} : \tau$ if $n = 0$\end{rep}.
The notation $t\typ{\tau}$ indicates that term $t$ has type~$\tau$.
\begin{rep}%
Functions invocations $\const{f}(t_1,\ldots,t_n)$
apply the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols, called constants, can appear without
parentheses in terms.
\end{rep}
The set $\Funcs$ must at least contain
$\const{true},\, \const{false} : \ty{bool}$, interpreted as truth values.
The only predicate is equality ($\teq$)\begin{rep} and belongs to the logical symbols\end{rep}.
Other predicates can be represented as functions to $\ty{bool}$,
with $\const{p}(\ldots)$ abbreviating $\const{p}(\ldots) \teq \const{true}$.
\begin{rep}The operator $\bigwedge_{\,i}\, \varphi_i$ abbeviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$. \end{rep}%
Finally, $[n] = \{1,\ldots, n\}$ and
$\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$.

@@

Each datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already known (co)datatypes, or among the (co)datatypes
being introduced.
%
To every argument corresponds a selector. The names for the (co)data\-types, the
constructors, and the selectors must be
\begin{conf}fresh\end{conf}\begin{rep}distinct and different from
existing names\end{rep}%
\dotReportFootnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by SMT-LIB 2, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
\!(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}= \smash{\const{C}_{11\!}(\bigl[\const{s}_{11\!}^1{:}\bigr]\vthinspace \tau_{11\!}^1, \ldots, \bigl[\const{s}_{11\!}^{n_{11\!}}{:}\bigr]\vthinspace \tau_{11\!}^{n_{11\!}})} \mid \cdots \mid \smash{\const{C}_{1m_1\!}(\ldots)} \\[-.5\jot]
   \smash{\vdots\,\,\,} \\[-1.5\jot]
  \keyw{and}\; \,\delta_{\elllx} & =\, \smash{\const{C}_{\elll 1\!}(\bigl[\const{s}_{\elll 1\!}^1{:}\bigr]\vthinspace \tau_{\elll 1\!}^1, \ldots, \bigl[\const{s}_{\elll 1\!}^{n_{\elll 1\!}}{:}\bigr]\vthinspace \tau_{\elll 1\!}^{n_{\elll 1\!}})} \mid \cdots \mid \smash{\const{C}_{\elll m_{\elllx}\!}(\ldots)}
\end{aligned}
\]
%
with
$\smash{\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,k_{\smash{i\negvthinspace j}}} \to \delta_i}$
and $\smash{\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}}$.
Defaults are assumed for the selector names if they are omitted.
The $\delta$ constructors and selectors are denoted by $\Ctr^\delta$ and
$\Sel^\delta$.

@@

Given a signature $\Sig$, we write $\thD$ to refer to the \emph{theory of
datatypes and codatatypes},
which in addition to $\Sig$ defines a class of $\Sig$-interpretations $\mathbf{J}$,
namely the ones that satisfy the %axioms --- existence of all infinite codatatype values cannot be axiomatized
properties mentioned in this section. In particular, the interpretations
in~$\mathbf{J}$ share the same fixed interpretation for constructor terms and correctly
applied selector terms but may differ on wrongly applied
selector terms and variables.
%%% @ANDY: Variables are normally not part of the interpretation, but given by
%%% an additional environment. Or is this done differently in the SMT community?
A formula $\varphi$ is $\thD$-\emph{satisfiable} if there
exists an interpretation in $\mathbf{J}$ that satisfies $\varphi$.
% otherwise, it is $\thD$-unsatisfiable.  (I trust the readers' intelligence when
% it comes to English prefixes like non-, in-, un-, and a-.)
\rem{more?}

@@

The mapping $\Val$ is computed as follows.
%
With each equivalence class $\ec{u}$, we associate a fresh variable $\Varec{u}$
of the same type as $u$
%not occurring in $\tEc$,
and set $\Val\ec{u} := {\Varec{u}}$. Therefore, there are initially no
constraints on the values for any equivalence class $\ec{u}$. The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the substitution of $t$ for $x$ in $\Val$'s range.
It is easy to see that the height of terms produced as a result of the unfolding
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

@@

First, we define the set of interpretations of the theory $\thD$,
which requires additional terminology concerning $\mu$-terms.
Given a $\mu$-term $t$ with subterm $u$,
%%% @ANDY: I'm taking the liberty of renaming "interpretation" to "expansion". Interpretation
%%% sounds semantic to me (although it is sometimes used for syntactic transformations).
the \emph{expansion of $u$ with respect to $t$} is the $\mu$-term $\interp{t}{u}^\emptyset$,
abbreviated to
$\interp{t}{u}$, as returned by the following recursive function:
%where $x \leadsto \MU x.\; \const{C}( \vec u ) \in t$ indicates that $\MU x.\;
%\const{C}( \vec u )$ is the subterm of $t$ that binds this occurrence of
%variable $x$:
\[\begin{array}{r@{}c@{}l}
\interp{t}{x}^B & {} \,=\, {} &
\textstyle\relax{\begin{cases}
    \MU x.\; \const{C}\smash{\bigl( \interp{t}{ \vec u}^{\smash{B \,\cup\, \{ x \}}}\bigr)} & \text{if this $x \notin B$ is
    bound by $\MU x.\; \const{C}( \vec u )$ in $t$}  \\[-\jot]
    x & \text{otherwise}
\end{cases}} ~ \\
\interp{t}{\MU x.\; \const{C}( \vec u )}^B & = &
\textstyle\begin{cases}
    \MU x.\; \const{C}\smash{\bigl( \interp{t}{ \vec u }^{\smash{B \,\cup\, \{ x \}}}\bigr)} & \text{if }x \not\in B \\[-\jot]
    x & \text{otherwise}
\end{cases}
\end{array}\]
%
\rem{TODO: add small examples}

The $\mu$-term $u$
is a \emph{self-similar subterm} of~$t$ if
$u$ is a proper subterm of $t$,
$t$ is of the form $\MU x.\; \const{C}( t_1, \ldots, t_n )$,
$u$ it is of the form $\MU y.\; \const{C}( u_1, \ldots, u_n )$,
and $\interp{t}{ t_k } \aequiv \interp{t}{ u_k }$ for all $k \in [n]$.
The term $t$ is \emph{normal} if it does not contain a self-similar subterm
and all of its proper subterms are also normal.
For example, $t = \MU x.\; \const{C}( \MU y.\; \const{C}( y ) )$ is abnormal
because $\MU y.\; \const{C}( y )$ is a self-similar subterm of $t$.
Their arguments have the same expansion with respect to~$t$:
$\interp{t}{\MU y.\; \const{C}( y )} =
\MU y.\; \const{C}\bigl( \interp{t}{y}^{\smash{\{y\}}} \bigr) =
\MU y.\; \const{C}( y )$
is $\alpha$-equivalent to
$\interp{t}{y} =
\MU y.\; \const{C}\bigl( \interp{t}{y}^{\smash{\{y\}}} \bigr) =
\MU y.\; \const{C}( y )$.
The term $t = \MU x.\; \const{C}( \MU y.\; \const{C}( x ) )$ is also abnormal,
since $\MU y.\; \const{C}( x )$ is a self-similar subterm of $t$,
noting that
%
\begin{align*}
\interp{t}{\MU y.\; \const{C}( x )}
& = \MU y.\; \const{C}\bigl( \interp{t}{x}^{\smash{\{y\}}} \bigr) \\
& = \MU y.\; \const{C}\bigl( \interp{t}{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) ) }^{\smash{\{y\}}} \bigr) \\
& = \MU y.\; \const{C}\bigl( \MU x.\; \const{C}( \interp{t}{\MU y.\; \const{C}( x )}^{\smash{\{x,y\}}} ) \bigr) \\
& = \MU y.\allowbreak\; \const{C}( \MU x.\allowbreak\; \const{C}( y ) )
\end{align*}
is $\alpha$-equivalent to $\interp{t}{x} = t$.
% for the renaming $\{ x \mapsto y, y \mapsto x \}$.

@@

We now show that $\Val = \nf{\Val}$.
Assume by contradition $\Val \ec{x} \neq \nf{\Val \ec{x}}$ for some $\Val \ec{x}$ of minimal size.
We have that $\Val \ec{x}$ is of the form $\MU {\Varec{x}}.\; \const{C}( t_1, \ldots, t_n )$.
Due to our construction of $\Val$,
we know $\ec{x}$ contains a constructor $\const{C}( z_1, \ldots, z_n )$ and $\muvar( t_i ) = z_i$ for $i \in [n]$.
Since $\Val \ec{x}$ is a minimal, it
contains a subterm of the form $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$
where $\interp{\Val \ec{x}}{ t_i } \aequiv \interp{\Val \ec{x} }{ u_i }$ for $i \in [n]$.
Due to our construction of $\Val$, $\ec{y}$ contains a constructor $\const{C}( w_1, \ldots, w_n )$ and $\muvar( u_i ) = w_i$ for $i \in [n]$.
Since \rn{Cong} does not apply to $\Ec$
we have $\ec{w_{\negvthinspace j}}$ and $\ec{z_{\negvthinspace j}}$ are distinct for some $j$.
By (2), $\interp{\Val \ec{x}}{ t_{\negvthinspace j} } = \Val \ec{z_{\negvthinspace j}}$
and $\interp{\Val \ec{x}}{ u_{\negvthinspace j} } = \Val \ec{w_{\negvthinspace j}}$,
which are not $\alpha$-equivalent by (3), contradicting the fact that $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$ is a self-similar subterm of $\Val \ec{x}$.
Thus, $\Val = \nf{\Val}$, and (2) and (3) hold for $\nf{\Val}$.

@@

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\M( t )$ is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) }$ for some fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \ValC \ec{ u_1 }, \ldots, \ValC \ec{ u_n } )}$.
Call this term $t_u$.
%By Lemma~\ref{lem:model-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{t}$ is a term
%of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
%where $\interp{\ValC \ec{t}}{w_i} = \ValC \ec{u_i}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$,
by our construction of $\ValC$ we have $\ValC \ec{t}$ is a term
of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
where for each $i \in [n]$,
$\muvar(w_i) = {\Varec{u_i}}$.  Thus by Lemma~\ref{lem:model-completion}\PART{2},
$\interp{\ValC \ec{t}}{w_i} = \ValC \ec{u_i}$.
For each $i \in [n]$, let $u_\iPrime$ be the $i$\vvthinspace th argument of $t_u$.
By Lemma~\ref{lem:mu-norm-arg}, $\interp{t_u}{u_\iPrime} \vsim \ValC \ec{u_i}$.
Thus $\interp{t_u}{u_\iPrime} \vsim \interp{\ValC \ec{t}}{ w_i }$. % for each $i \in [n]$.
%By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Thus, $\M( t ) \vsim t_u \vsim \ValC \ec{t}$, and
we have $\M( t ) \vsim \ValC \ec{t}$.

If $t$ is a selector term of the form $\const s^k_{\negvthinspace j}( u )$,
since \rn{Split} does not apply to $\Fc$,
$\ec{u}$ must contain a term of the form $\smash{\const C_{\negvthinspace\jPrime}\bigl( \const s^1_{\negvthinspace\jPrime}( u ), \ldots, \const s^{n}_{\negvthinspace\jPrime}( u ) \bigr)}$ for some $\jPrime$.
%By Lemma~\ref{lem:model-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{u}$ is of the form
%$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
%where $\interp{\ValC \ec{u} }{ w_i } = \ValC \ec{ \const s^i_{\negvthinspace\jPrime}( u )}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$,
by construction $\ValC$ must be of the form
$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
where for each $i \in [n]$,
$\muvar(w_i) = \smash{\Varec{\vphantom{I}\smash{\const s^i_{\negvthinspace\jPrime}( u )}}}$ and thus by Lemma~\ref{lem:model-completion}\PART{2},
$\interp{\ValC \ec{u}}{ w_i } = \ValC \ec{ \smash{\const s^i_{\negvthinspace\jPrime}( u )}}$.
If $j = \jPrime$, then $\M( t )$ is $\alpha$-equivalent with $\interp{\ValC \ec{u}}{w_k}$, which is equal to $\ValC \ec{ \smash{ \const s^k_{\negvthinspace j}( u )}}$,
which is $\ValC \ec{t}$.
If $j \neq \jPrime$, since \rn{Cong} does not apply,
$\smash{\const s^k_{\negvthinspace j}( u )}$ and $\smash{\const s^k_{\negvthinspace j}( u' )}$
are in distinct equivalence classes only if $u$ and $u'$ are in distinct equivalence classes.
Due to the induction hypothesis and Lemma~\ref{lem:model-completion}\PART{3}, $\M( u ) \neq \M( u' )$ for all such $u$, $u'$.
Thus, we may interpret $\smash{\M(\const s^k_{\negvthinspace j})( \M( u ) )}$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\ValC \ec{ t }$.

@@

\paragraph{Integration into {\relax\XXXL($T$)}.}
The calculus is implemented as a \emph{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a input set of
clauses $F$ in CNF, the \XXXL($T$) procedure
incrementally builds partial assignments from the atomic formulas of $F$ to
truth values such that no clause in $F$ is falsified. We may interpret such a
partial assignment as a set of literals where $a \in M$ for all atoms $a$ that
it assigned the value $\top$, and $\neg\; a \in M$ for all atoms that it assigns
the value $\bot$. Then, at a high level, by the Nelson--Oppen combination
framework (its variant in CVC4 is described in~\cite{jovanovic2011sharing}), 
each $T_i$-solver for $i = 1, \ldots, l$ takes as input a combination
of (1) the purified form of $T_i$-constraints occurring in $M$, where terms
containing symbols not belonging to $T_i$ are replaced by fresh variables of the
same type; (2) additional equalities and disequalities between variables of
types not belonging to $T_i$. Let us call this set $M_i$. Each $T_i$ solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\; C$ is added to $F$, adds a clause to $F$, or does nothing.
When $M$ is a complete assignment for $F$, a theory solver can only choose to do
nothing when $M_i$ is indeed $T_i$-satisfiable.

>>>>

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$).
\end{rep}
Names starting with an uppercase letter \begin{rep}(e.g.,
$\const{S}$) \end{rep}are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (for $n \ge 0$)
and a return type $\tau$, written
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$\begin{rep},
which collapses to $\const{f} : \tau$ if $n = 0$\end{rep}.
The notation $t\typ{\tau}$ indicates that term $t$ has type~$\tau$.
\begin{rep}%
Functions invocations $\const{f}(t_1,\ldots,t_n)$
apply the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols, called constants, can appear without
parentheses in terms.
\end{rep}
The set $\Funcs$ must at least contain
$\const{true},\, \const{false} : \ty{bool}$, interpreted as truth values.
The only predicate is equality ($\teq$)\begin{rep} and belongs to the logical symbols\end{rep}.
Other predicates can be represented as functions to $\ty{bool}$,
with $\const{p}(\ldots)$ abbreviating $\const{p}(\ldots) \teq \const{true}$.
\begin{rep}The operator $\bigwedge_{\,i}\, \varphi_i$ abbeviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$. \end{rep}%
Finally, $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$.

@@

Each (co)datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already known (co)datatypes, or among the (co)datatypes
being introduced.
%
To every argument corresponds a selector. The names for the (co)data\-types, the
constructors, and the selectors must be
\begin{conf}fresh\end{conf}\begin{rep}distinct and different from
existing names\end{rep}%
\dotReportFootnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by SMT-LIB 2, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
\!(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}= \smash{\const{C}_{11\!}(\bigl[\const{s}_{11\!}^1{:}\bigr]\vthinspace \tau_{11\!}^1, \ldots, \bigl[\const{s}_{11\!}^{n_{11\!}}{:}\bigr]\vthinspace \tau_{11\!}^{n_{11\!}})} \mid \cdots \mid \smash{\const{C}_{1m_1\!}(\ldots)} \\[-.5\jot]
   \smash{\vdots\,\,\,} \\[-1.5\jot]
  \keyw{and}\; \,\delta_{\elllx} & =\, \smash{\const{C}_{\elll 1\!}(\bigl[\const{s}_{\elll 1\!}^1{:}\bigr]\vthinspace \tau_{\elll 1\!}^1, \ldots, \bigl[\const{s}_{\elll 1\!}^{n_{\elll 1\!}}{:}\bigr]\vthinspace \tau_{\elll 1\!}^{n_{\elll 1\!}})} \mid \cdots \mid \smash{\const{C}_{\elll m_{\elllx}\!}(\ldots)}
\end{aligned}
\]
%
with
$\smash{\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,k_{\smash{i\negvthinspace j}}} \to \delta_i}$
and $\smash{\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}}$.
Defaults are assumed for the selector names if they are omitted.
The $\delta$ constructors and selectors are denoted by $\Ctr^\delta$ and
$\Sel^\delta$.

@@

Given a signature $\Sig$, we write $\thD$ to refer to the \emph{theory of
datatypes and codatatypes},
which in addition to $\Sig$ defines a class of $\Sig$-interpretations $\mathbf{J}$,
namely the ones that satisfy the %axioms --- existence of all infinite codatatype values cannot be axiomatized
properties mentioned in this section, including (co)induction. In particular, the interpretations
in~$\mathbf{J}$ share the same fixed interpretation for constructor terms and correctly
applied selector terms but may differ on wrongly applied
selector terms and variables.
%%% @ANDY: Variables are normally not part of the interpretation, but given by
%%% an additional environment. Or is this done differently in the SMT community?
A formula $\varphi$ is $\thD$-\emph{satisfiable} if there
exists an interpretation in $\mathbf{J}$ that satisfies $\varphi$.
% otherwise, it is $\thD$-unsatisfiable.  (I trust the readers' intelligence when
% it comes to English prefixes like non-, in-, un-, and a-.)
\rem{more?}

@@

The mapping $\Val$ is computed as follows.
%
With each equivalence class $\ec{u}$, we associate a fresh variable $\Varec{u}$
of the same type as $u$
%not occurring in $\tEc$,
and set $\Val\ec{u} := {\Varec{u}}$. Therefore, there are initially no
constraints on the values for any equivalence class $\ec{u}$. The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the variable-capturing substitution of $t$ for
$x$ in $\Val$'s range. It is easy to see that the height of terms produced as a
result of the unfolding is bounded by the number of equivalence classes of
$\Ec$, and thus the construction of $\Val$ will terminate.

@@

First, we define the set of interpretations of the theory $\thD$,
which requires additional terminology concerning $\mu$-terms.
Given a $\mu$-term $t$ with subterm $u$,
%%% @ANDY: I'm taking the liberty of renaming "interpretation" to "expansion". Interpretation
%%% sounds semantic to me (although it is sometimes used for syntactic transformations).
the \emph{expansion of $u$ with respect to $t$} is the $\mu$-term $\interp{t}{u}^\emptyset$,
abbreviated to
$\interp{t}{u}$, as returned by the following recursive function:
%where $x \leadsto \MU x.\; \const{C}( \vec u ) \in t$ indicates that $\MU x.\;
%\const{C}( \vec u )$ is the subterm of $t$ that binds this occurrence of
%variable $x$:
\[\begin{array}{r@{}c@{}l}
\interp{t}{x}^B & {} \,=\, {} &
\textstyle\relax{\begin{cases}
    \MU x.\; \const{C}\smash{\bigl( \interp{t}{ \vec u}^{\smash{B \,\cup\, \{ x \}}}\bigr)} & \text{if this $x \notin B$ is
    bound by $\MU x.\; \const{C}( \vec u )$ in $t$}  \\[-\jot]
    x & \text{otherwise}
\end{cases}} ~ \\
\interp{t}{\MU x.\; \const{C}( \vec u )}^B & = &
\textstyle\begin{cases}
    \MU x.\; \const{C}\smash{\bigl( \interp{t}{ \vec u }^{\smash{B \,\cup\, \{ x \}}}\bigr)} & \text{if }x \not\in B \\[-\jot]
    x & \text{otherwise}
\end{cases}
\end{array}\]
%
\rem{TODO: add small examples}
\rem{TODO: explain why terminates}

The $\mu$-term $u$
is a \emph{self-similar subterm} of~$t$ if
$u$ is a proper subterm of $t$,
$t$ is of the form $\MU x.\; \const{C}( t_1, \ldots, t_n )$,
$u$ is of the form $\MU y.\; \const{C}( u_1, \ldots, u_n )$,
and $\interp{t}{ t_k } \aequiv \interp{t}{ u_k }$ for all $k$. % \in [n]
The term $t$ is \emph{normal} if it does not contain a self-similar subterm
and all of its proper subterms are also normal.
For example, $t = \MU x.\; \const{C}( \MU y.\; \const{C}( y ) )$ is abnormal
because $\MU y.\; \const{C}( y )$ is a self-similar subterm of $t$.
Their arguments have the same expansion with respect to~$t$:
$\interp{t}{\MU y.\; \const{C}( y )} =
\MU y.\; \const{C}\bigl( \interp{t}{y}^{\smash{\{y\}}} \bigr) =
\MU y.\; \const{C}( y )$
is $\alpha$-equivalent to
$\interp{t}{y} =
\MU y.\; \const{C}\bigl( \interp{t}{y}^{\smash{\{y\}}} \bigr) =
\MU y.\; \const{C}( y )$.
The term $t = \MU x.\; \const{C}( \MU y.\; \const{C}( x ) )$ is also abnormal,
since $\MU y.\; \const{C}( x )$ is a self-similar subterm of $t$,
noting that
%
\begin{align*}
\interp{t}{\MU y.\; \const{C}( x )}
& = \MU y.\; \const{C}\bigl( \interp{t}{x}^{\smash{\{y\}}} \bigr) \\
& = \MU y.\; \const{C}\bigl( \interp{t}{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) ) }^{\smash{\{y\}}} \bigr) \\
& = \MU y.\; \const{C}\bigl( \MU x.\; \const{C}( \interp{t}{\MU y.\; \const{C}( x )}^{\smash{\{x,y\}}} ) \bigr) \\
& = \MU y.\allowbreak\; \const{C}( \MU x.\allowbreak\; \const{C}( y ) )
\end{align*}
is $\alpha$-equivalent to $\interp{t}{x} = t$.
% for the renaming $\{ x \mapsto y, y \mapsto x \}$.

@@

We now show that $\Val = \nf{\Val}$.
Assume by contradition $\Val \ec{x} \neq \nf{\Val \ec{x}}$ for some $\Val \ec{x}$ of minimal size.
We have that $\Val \ec{x}$ is of the form $\MU {\Varec{x}}.\; \const{C}( t_1, \ldots, t_n )$.
Due to our construction of $\Val$,
we know $\ec{x}$ contains a constructor $\const{C}( z_1, \ldots, z_n )$ and $\muvar( t_i ) = z_i$ for some $i$ % \in [n]$.
Since $\Val \ec{x}$ is a minimal, it
contains a subterm of the form $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$
where $\interp{\Val \ec{x}}{ t_i } \aequiv \interp{\Val \ec{x} }{ u_i }$ for some $i$. % \in [n]$.
Due to our construction of $\Val$, $\ec{y}$ contains a constructor $\const{C}( w_1, \ldots, w_n )$ and $\muvar( u_i ) = w_i$ for some $i$. % \in [n]$.
Since \rn{Cong} does not apply to $\Ec$
we have $\ec{w_{\negvthinspace j}}$ and $\ec{z_{\negvthinspace j}}$ are distinct for some $j$.
By (2), $\interp{\Val \ec{x}}{ t_{\negvthinspace j} } = \Val \ec{z_{\negvthinspace j}}$
and $\interp{\Val \ec{x}}{ u_{\negvthinspace j} } = \Val \ec{w_{\negvthinspace j}}$,
which are not $\alpha$-equivalent by (3), contradicting the fact that $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$ is a self-similar subterm of $\Val \ec{x}$.
Thus, $\Val = \nf{\Val}$, and (2) and (3) hold for $\nf{\Val}$.

@@

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\M( t )$ is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) }$ for some fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \ValC \ec{ u_1 }, \ldots, \ValC \ec{ u_n } )}$.
Call this term $t_u$.
%By Lemma~\ref{lem:model-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{t}$ is a term
%of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
%where $\interp{\ValC \ec{t}}{w_i} = \ValC \ec{u_i}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$,
by our construction of $\ValC$ we have $\ValC \ec{t}$ is a term
of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
where for each $i \in \{1,\ldots, n\}$, % \in [n]$,
$\muvar(w_i) = {\Varec{u_i}}$.  Thus by Lemma~\ref{lem:model-completion}\PART{2},
$\interp{\ValC \ec{t}}{w_i} = \ValC \ec{u_i}$.
For each $i \in \{1,\ldots, n\}$, let $u_\iPrime$ be the $i$\vvthinspace th argument of $t_u$.
By Lemma~\ref{lem:mu-norm-arg}, $\interp{t_u}{u_\iPrime} \vsim \ValC \ec{u_i}$.
Thus $\interp{t_u}{u_\iPrime} \vsim \interp{\ValC \ec{t}}{ w_i }$. % for each $i \in [n]$.
%By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Thus, $\M( t ) \vsim t_u \vsim \ValC \ec{t}$, and
we have $\M( t ) \vsim \ValC \ec{t}$.

If $t$ is a selector term of the form $\const s^k_{\negvthinspace j}( u )$,
since \rn{Split} does not apply to $\Fc$,
$\ec{u}$ must contain a term of the form $\smash{\const C_{\negvthinspace\jPrime}\bigl( \const s^1_{\negvthinspace\jPrime}( u ), \ldots, \const s^{n}_{\negvthinspace\jPrime}( u ) \bigr)}$ for some $\jPrime$.
%By Lemma~\ref{lem:model-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{u}$ is of the form
%$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
%where $\interp{\ValC \ec{u} }{ w_i } = \ValC \ec{ \const s^i_{\negvthinspace\jPrime}( u )}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$,
by construction $\ValC$ must be of the form
$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
where for each $i \in \{1,\ldots, n\}$,
$\muvar(w_i) = \smash{\Varec{\vphantom{I}\smash{\const s^i_{\negvthinspace\jPrime}( u )}}}$ and thus by Lemma~\ref{lem:model-completion}\PART{2},
$\interp{\ValC \ec{u}}{ w_i } = \ValC \ec{ \smash{\const s^i_{\negvthinspace\jPrime}( u )}}$.
If $j = \jPrime$, then $\M( t )$ is $\alpha$-equivalent with $\interp{\ValC \ec{u}}{w_k}$, which is equal to $\ValC \ec{ \smash{ \const s^k_{\negvthinspace j}( u )}}$,
which is $\ValC \ec{t}$.
If $j \neq \jPrime$, since \rn{Cong} does not apply,
$\smash{\const s^k_{\negvthinspace j}( u )}$ and $\smash{\const s^k_{\negvthinspace j}( u' )}$
are in distinct equivalence classes only if $u$ and $u'$ are in distinct equivalence classes.
Due to the induction hypothesis and Lemma~\ref{lem:model-completion}\PART{3}, $\M( u ) \neq \M( u' )$ for all such $u$, $u'$.
Thus, we may interpret $\smash{\M(\const s^k_{\negvthinspace j})( \M( u ) )}$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\ValC \ec{ t }$.

@@

\paragraph{Integration into {\relax\XXXL($T$)}.}
The calculus is implemented as a \emph{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and an input set of
clauses $F$ in CNF, the \XXXL($T$) procedure
incrementally builds partial assignments from the atomic formulas of $F$ to
truth values such that no clause in $F$ is falsified. We may interpret such a
partial assignment as a set of literals where $a \in M$ for all atoms $a$ that
it assigned the value $\top$, and $\neg\; a \in M$ for all atoms that it assigns
the value $\bot$. Then, at a high level, by the Nelson--Oppen combination
framework (its variant in CVC4 is described in~\cite{jovanovic2011sharing}), 
each $T_i$-solver for $i = 1, \ldots, l$ takes as input a combination
of (1) the purified form of $T_i$-constraints occurring in $M$, where terms
containing symbols not belonging to $T_i$ are replaced by fresh variables of the
same type; (2) additional equalities and disequalities between variables of
types not belonging to $T_i$. Let us call this set $M_i$. Each $T_i$ solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\; C$ is added to $F$, adds a clause to $F$, or does nothing.
When $M$ is a complete assignment for $F$, a theory solver can only choose to do
nothing when $M_i$ is indeed $T_i$-satisfiable.
